---
title: "Week 5 Lab: Timie series analysis and a map!"
author: "Anne-Marie Parkinson"
date: "February 6, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F)
```

```{r}
# load packages

library(tidyverse)
library(janitor)
library(lubridate)
library(here)
library(paletteer)
library(tsibble)
library(fable)
library(fabletools)
library(feasts)
library(forecast)
library(sf)
library(tmap)
library(mapview)

```

## Monthly US energy consumption (renewables)

```{r}

#load data

us_renew <- read_csv(here("data", "renewables_cons_prod.csv")) %>% 
  clean_names()
  
# sub-data: make descriptions lowercase, only keep observations including the phrase "consumption", remove observations in the description column with the phrase "total"
  
renew_clean <- us_renew %>% 
  mutate (description = str_to_lower(description)) %>% 
  filter(str_detect(description, pattern = "consumption")) %>% #only keeps observations in this column with the word consumption
  filter(!str_detect(description, patter = "total")) # gets rid of any obs where the word total is incldued in the observation. use the ! in front of str_detect to remove (instead of keep) observations

```





```{r}
# convert "yyyymm" column to a date

renew_date <- renew_clean %>% 
  mutate(yr_mo_day = lubridate::parse_date_time(yyyymm, "ym"))
# get NAs bc some observations in the original yyyymm column are not in the traditional yyyymm format (ex/ 191913, theres no 13th month). Gives all values the day value of 01

renew_date <- renew_date %>% 
  mutate(month_sep = yearmonth(yr_mo_day)) %>% # provides the yyyymm data in "yyyy first 3 letter of the Name of Month" format
  mutate(value = as.numeric(value)) %>% 
  drop_na(month_sep, value)

#make a data frame where the month and dyear are in separate columns

renew_parsed <- renew_date %>% 
  mutate(year = year(yr_mo_day)) %>% 
  mutate(month = month(yr_mo_day, label = T))# to keep the month stored as a labeled name (in 3 letter abrev. frmat) instead of a number (which is the default) use label = T

  

```

## Visually explore data

```{r}

renew_gg <- ggplot(data = renew_date, aes(x = month_sep, y = value)) +
  geom_line() #shows useless, random graph. Need to group the data by a specific column

renew_gg <- ggplot(data = renew_date, aes(x = month_sep, y = value, group = description)) +
  geom_line(aes(color = description)) #group = redundant when base color on an aesthetic, but can still keep it 

# use view(palettes_d_names) in the console to get a datasheet of the diff packages and the names and how many colors in the package (this is in the length column). Need to use palettes_d bc have aes based on discrete data. use palettes_c_names for data for continuous data

```

##add colors to graph with paletteer

```{r}
renew_gg +
  scale_color_paletteer_d("calecopal::figmtn") # format= (package::palette name). make sure to use ..._d bc the color is based on discrete data. Make sure palette you use has enough colors for the number of discrete classes you have. in this case, we need a package with 7+ colors
```


## coerce data renew_parse to a tsibble
```{r}

renew_ts <- as_tsibble(renew_parsed, key = description, index = month_sep)

#the yyyy Mon is a tibble freindly format. can also use other date formats

```


```{r}
renew_ts %>% autoplot(value)
renew_ts %>% gg_subseries(value) #diff consumption sources split up by month
# renew_ts %>% gg_season(value) --> problem with source package

ggplot(data = renew_parsed, aes(x = month, y = value, group =year)) +
  geom_line(aes(color = year)) +
  facet_wrap(~description,
             ncol = 1,
             scales = "free", 
             strip.position = "right") # this puts the name of the groups on the side instead of on the top, which is the default

```


##just look at hydroelectric energy consumption

```{r}
hydro_ts <- renew_ts %>% 
  filter(description == "hydroelectric power consumption")

hydro_ts %>% autoplot(value)
hydro_ts %>%  gg_subseries(value) #look at change over each month
ggplot(data = hydro_ts, aes(x = month, y = value, group =year)) +
  geom_line(aes(color = year))

```

## quarterly average consumption for hydro

```{r}
hydro_quarterly <- hydro_ts %>% 
  index_by(year_qu = ~(yearquarter(.))) %>%#why put a period here? how does r know which column to index by? also what are the quarters? jan-mar, apr-june, etc?
  summarise(avg_consumption = mean(value))

```

##Decompose the hydro_ts data frame

```{r}
#using moving average

#create the model
dcmp <- hydro_ts %>% 
  model(STL(value ~ season(window = 5))) # says decompose the values in the value column using a window of 5

#plot deconstructed model results 
components(dcmp) %>% autoplot() # can manually add equal scales like would in ggplot

```


```{r}
hydro_ts %>% 
  ACF(value) %>% 
  autoplot

```
confirms there is a seasonality trend 

#DANGER: dont copy this method of forecasting. its just a quick example!!!!

```{r}
hydro_model <- hydro_ts %>% 
  model(
    ARIMA(value)
  ) %>% 
  fabletools::forecast(h = "4 years") #the package recognizes the saying 4 years!

hydro_model %>% autoplot() # only shows the forecasted values/years 

hydro_model %>% autoplot(hydro_ts) #plots all data from hydro_ts including the predicted values

hydro_model %>% autoplot(filter(hydro_ts, year(month_sep) > 2010)) # in hydro_ts dataset, only plot just the years after 2010 plus the predicted values

```

## same as above but with 2 diff model typesforecasting the same values. typically papers will publish more than one modell bc we cant say for sure if that one prediction will happen. 
```{r}
hydro_model <- hydro_ts %>% 
  model(
    ARIMA(value),
    ETS(value)
  ) %>% 
  fabletools::forecast(h = "4 years") #the package recognizes the saying 4 years!

hydro_model %>% autoplot() # only shows the forecasted values/years 

hydro_model %>% autoplot(hydro_ts) #plots all data from hydro_ts including the predicted values

hydro_model %>% autoplot(filter(hydro_ts, year(month_sep) > 2010)) # in hydro_ts dataset, only plot just the years after 2010 plus the predicted values

```



##World map with SF

```{r}
world <- read_sf(dsn = here("data", "TM_WORLD_BORDERS_SIMPL-0.3-1"), 
                 layer = "TM_WORLD_BORDERS_SIMPL-0.3")
```

```{r}
mapview(world)
```











